---
title: "Initial data exploration"
author: "Abby Lewis"
date: "2024-08-21"
output: html_document
---

Load data and packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)

ch4 <- read_csv("Raw_data/GENX_Flux_SD_Loggernet_2021-05-05_to_2022-12-31_derived.csv", show_col_types = FALSE) %>%
  mutate(Temp = sub("e", "", zone_treatment),
         Temp = ifelse(Temp == "amb", 0, Temp),
         Temp = as.numeric(Temp))
```

Initial plots

```{r}
ch4 %>%
  filter(CH4_rsquared > 0.9) %>%
  ggplot(aes(x = date, y = CH4_slope_ppm_per_day)) +
  geom_line(aes(group = plot_id))+
  facet_wrap(~zone_treatment)

ch4 %>%
  filter(CH4_rsquared > 0.9) %>%
  ggplot(aes(x = date, y = CH4_slope_ppm_per_day, color = zone_treatment)) +
  geom_smooth(se = F) +
  scale_color_viridis_d()
```

Something is interesting about how temperature treatments affect variability of methane fluxes. There is greater day-to-day variability in the ambient treatment compared to the warmed treatment, particularly when CH4 fluxes are high.

```{r}
all_dates <- ch4 %>%
  mutate(date_num = as.numeric(as.Date(flux_timestamp)),
         Year = year(flux_timestamp)) %>%
  group_by(plot_id, chamber_treatment, zone_treatment, Year) %>%
  expand(date_num = seq(min(date_num), max(date_num))) %>%
  mutate(Temp = sub("e", "", zone_treatment),
         Temp = ifelse(Temp == "amb", "0", Temp),
         Temp = as.numeric(Temp))

rolling <- ch4 %>% 
  filter(CH4_rsquared > 0.9) %>%
  mutate(date_num = as.numeric(as.Date(flux_timestamp))) %>%
  group_by(plot_id, chamber_treatment, zone_treatment) %>% #, date_num) %>%
  #summarize(CH4_slope_ppm_per_day = mean(CH4_slope_ppm_per_day, na.rm = T)) %>%
  full_join(all_dates) %>%
  arrange(date_num) %>%
  mutate(Rolling_sd = slider::slide_index_dbl(.x = CH4_slope_ppm_per_day, 
                                                  .i = date_num, 
                                                  .f = sd, 
                                                  na.rm = T,
                                                  .before = 30),
         Rolling_mean = slider::slide_index_dbl(.x = CH4_slope_ppm_per_day, 
                                                  .i = date_num, 
                                                  .f = mean, 
                                                  na.rm = T,
                                                  .before = 30),
         Temp_group = ifelse(Temp >= 3, "High", "Low")) 

rolling %>%
  pivot_longer(cols = c("Rolling_sd", "Rolling_mean"), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = as.Date(date_num), y = Value, color = zone_treatment, group = chamber_treatment)) +
  geom_line()+
  facet_grid(rows = vars(Temp_group), cols = vars(Variable)) +
  scale_color_viridis_d()

rolling %>%
  ggplot(aes(x = Rolling_mean, y = Rolling_sd, color = zone_treatment)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  scale_color_viridis_d()
```

What if we use medians instead?

```{r}
all_dates <- ch4 %>%
  mutate(date_num = as.numeric(as.Date(flux_timestamp)),
         Year = year(flux_timestamp)) %>%
  group_by(plot_id, chamber_treatment, zone_treatment, Year) %>%
  expand(date_num = seq(min(date_num), max(date_num))) %>%
  mutate(Temp = sub("e", "", zone_treatment),
         Temp = ifelse(Temp == "amb", "0", Temp),
         Temp = as.numeric(Temp))

rolling <- ch4 %>% 
  filter(CH4_rsquared > 0.9) %>%
  mutate(date_num = as.numeric(as.Date(flux_timestamp))) %>%
  group_by(plot_id, chamber_treatment, zone_treatment) %>% #, date_num) %>%
  #summarize(CH4_slope_ppm_per_day = mean(CH4_slope_ppm_per_day, na.rm = T)) %>%
  full_join(all_dates) %>%
  arrange(date_num) %>%
  mutate(Rolling_iqr = slider::slide_index_dbl(.x = CH4_slope_ppm_per_day, 
                                                  .i = date_num, 
                                                  .f = IQR, 
                                                  na.rm = T,
                                                  .before = 30),
         Rolling_median = slider::slide_index_dbl(.x = CH4_slope_ppm_per_day, 
                                                  .i = date_num, 
                                                  .f = median, 
                                                  na.rm = T,
                                                  .before = 30),
         Temp_group = ifelse(Temp >= 3, "High", "Low")) 

rolling %>%
  pivot_longer(cols = c("Rolling_iqr", "Rolling_median"), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = as.Date(date_num), y = Value, color = zone_treatment, group = chamber_treatment)) +
  geom_line()+
  facet_grid(rows = vars(Temp_group), cols = vars(Variable)) +
  scale_color_viridis_d()

rolling %>%
  ggplot(aes(x = Rolling_median, y = Rolling_iqr, color = zone_treatment)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  scale_color_viridis_d() 
```

Is variability within a day also lower in high-temperature treatments?
Yes, when calculated as mean ± SD. Not when calculated as median ± IQR.

```{r}
daily <- ch4 %>% 
  filter(CH4_rsquared > 0.9) %>%
  mutate(date_num = as.numeric(as.Date(flux_timestamp))) %>%
  group_by(plot_id, chamber_treatment, zone_treatment, date_num) %>%
  mutate(Daily_sd = sd(CH4_slope_ppm_per_day, na.rm = T),
         Daily_mean = mean(CH4_slope_ppm_per_day, na.rm = T),
         Temp_group = ifelse(Temp >= 3, "High", "Low"),
         n = sum(!is.na(CH4_slope_ppm_per_day))) %>%
  filter(n > 5)

#Mean
daily %>%
  ggplot(aes(x = Daily_mean, y = Daily_sd, color = zone_treatment)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  scale_color_viridis_d()

daily <- ch4 %>% 
  filter(CH4_rsquared > 0.9) %>%
  mutate(date_num = as.numeric(as.Date(flux_timestamp))) %>%
  group_by(plot_id, chamber_treatment, zone_treatment, date_num) %>%
  mutate(Daily_iqr = IQR(CH4_slope_ppm_per_day, na.rm = T),
         Daily_median = median(CH4_slope_ppm_per_day, na.rm = T),
         Temp_group = ifelse(Temp >= 3, "High", "Low"),
         n = sum(!is.na(CH4_slope_ppm_per_day))) %>%
  filter(n > 5)

#Median
daily %>%
  ggplot(aes(x = Daily_median, y = Daily_iqr, color = zone_treatment)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  scale_color_viridis_d()
```

Is this true within a day?

```{r}
for_plot <- ch4 %>% 
  filter(CH4_rsquared > 0.9) %>%
  mutate(date_num = as.numeric(as.Date(flux_timestamp)),
         hour = hour(flux_timestamp),
         year = year(flux_timestamp),
         time = ifelse(hour < 17 & hour > 9, "day",
                       ifelse(hour > 20 | hour < 4, "night", 
                              "other"))) %>%
  filter(!time == "other") %>%
  group_by(zone_treatment, year, time, date_num) %>%
  summarize(mean = mean(CH4_slope_umol_per_day, na.rm = T)) 

#In 2021, all treatments were higher at night. In 2022, these differences are gone. No clear indication that treatment influences the magnitude of variability within a day
for_plot %>%
  ggplot(aes(x = time, y = mean)) +
  geom_hline(yintercept = 0, color = "grey")+
  geom_boxplot(aes(color = zone_treatment))+
  facet_grid(cols = vars(zone_treatment), rows = vars(year), scales = "free") +
  scale_color_viridis_d()+
  theme_bw()+
  coord_cartesian(ylim = c(-50, 100))+
  theme(legend.position = "bottom")

for_plot %>%
  pivot_wider(names_from = time, values_from = mean) %>%
  filter(!is.na(day), !is.na(night)) %>%
  ggplot(aes(x = day, y = night)) +
  geom_abline(intercept = 0, slope = 1, color = "grey") +
  geom_smooth(aes(color = zone_treatment), method = "lm")+
  facet_wrap(~year)+
  scale_color_viridis_d()
```

```{r}
drivers <- read_csv("Raw_data/GENX_SD_Export_Waterlevel_2021-05-06_to_2022-12-30_derived_clean_btemps.csv", 
                    show_col_types = F)

p <- drivers %>%
  filter(time2 > "2021-06-20", time2 < "2021-06-25") %>%
  ggplot(aes(x = time2, y = corrected_depth))+
  geom_line()

library(rtide)
library(scales)
data <- rtide::tide_height(
  "Annapolis (US Naval Academy), Severn River, Maryland",
  from = as.Date("2021-07-01"), to = as.Date("2021-08-01"),
  minutes = 10L, tz = "UTC"
)

data %>%
  ggplot(aes(x = DateTime, y = TideHeight)) +
  geom_line() +
  scale_y_continuous(name = "Tide Height (m)") +
  geom_line(data = drivers %>% filter(time2 > "2021-07-01", time2 < "2021-08-01"), aes(x = time2, y = corrected_depth/100), color = "red")
p

plotly::ggplotly(p)
```


